---
title: "Simulating clinical trials"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
pctl_list <- c(1, 25, 50, 75, 99)
N <- 350
T <- 1095
S <- 0.5
dat_prior <- list(N=N, T=T, S=S)
pts <- unlist(read.csv("2-paths-count.csv"))
dat_updated <- list(N=350, T=3*365, S=0.5, n=pts, t=length(pts))
```

## Paths in a simulation

When you simulate a clinical trial, you won't do this two times. You'll do it several thousand times, because computer cycles are cheap. Here's what a few thousand simulations might look like.

### Figure 2.1. 1000 simulations run prior to start of a clinical trial.

```{r paths-before-trial, echo=FALSE, fig.width=7, fig.height=4}
# For some reason, Stan is very slow for this simulation. I still want to
# use Stan though because it serves as a good springboard for many of the
# later simulations.
f <- "2-paths-before-trial.stan"
fit_tr1 <- stan(file=f,
  data=dat_prior, iter= 1000, warmup=100, chains = 4)

fit_tr1              %>%
  as.data.frame                %>%
  select(starts_with('Nstar')) -> sim_tr1
sim_tr1$j <- 1:dim(sim_tr1)[1]
sim_tr1 %<>% gather(x, y, starts_with('Nstar'))

sim_tr1$x %<>%
  sub("Nstar\\[", "", .) %>%
  sub("\\]", "", .)      %>%
  as.numeric

sim_tr1                                         %>%
  filter(x==1095)                               %>%
  use_series(y)                                 %>%
  quantile(pctl_list/100)                       -> y_ticks
y_ticks                                         %>%
  round                                         %>%
  paste0(" (", pctl_list, "%)")                 -> y_labels

sim_tr1 %>%
  ggplot(aes(x, y, group=j))                     +
  geom_step(alpha=0.01)                          +
  scale_y_continuous(breaks=y_ticks,
                     labels=y_labels,
                     position="right",
                     minor_breaks=NULL)          +
  scale_x_continuous(breaks=365*(0:3))           +
  xlab("Time (days)")                            +
  ylab("Estimated number of patients (Prior)")   +
  expand_limits(y=500)
```

I also want to talk about simulating a clinical trial during the trial. What this means is that you have information sample size up to a certain point in time and you want to simulate how rapidly the sample size might increase for the remainder of time in the trial.

### Figure 2.2. 1000 simulations run during a clinical trial.

```{r path-during-trial, echo=FALSE, fig.width=7, fig.height=4}
f <- "2-paths-during-trial.stan"
fit_tr2 <- stan(file=f,
  data=dat_updated, iter= 1000, warmup=100, chains = 4)

fit_tr2              %>%
  as.data.frame                %>%
  select(starts_with('Nstar')) -> sim_tr2
sim_tr2$j <- 1:dim(sim_tr2)[1]
sim_tr2 %<>% gather(x, y, starts_with('Nstar'))

sim_tr2$x %<>%
  sub("Nstar\\[", "", .) %>%
  sub("\\]", "", .)      %>%
  as.numeric

sim_tr2                                         %>%
  filter(x==T)                                  %>%
  use_series(y)                                 %>%
  quantile(pctl_list/100)                       -> y_ticks
y_ticks                                         %>%
  round                                         %>%
  paste0(" (", pctl_list, "%)")                 -> y_labels

sim_tr2 %>%
  ggplot(aes(x, y, group=j))                     +
  geom_step(alpha=0.01)                          +
  xlab("Time (days)")                            +
  ylab("Estimated number of patients (Updated)") +
  scale_y_continuous(breaks=y_ticks,
                     labels=y_labels,
                     position="right",
                     minor_breaks=NULL)          +
  scale_x_continuous(breaks=365*(0:3))           +
  expand_limits(y=500)                           

```

```{r save-everything}
save.image("2-paths.RData")
```
