---
title: "Simulating clinical trials"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
s_list <- c(0.1, 0.2, 0.5)
n_s <- length(s_list)
n_reps <- 1000
pctl_list <- c(1, 25, 50, 75, 99)
N <- 350
T <- 1095
dat_prior <- list(N=N, T=T, S=0.5, n=0, t=0)

```

## Gamma Poisson models

From this point onward, we'll summarize the simulations using boxplots and scatterplots, but it's worth remembering that what we are actually simulating is random paths.

Let's also focus on simulations of the total sample size. In a Bayesian perspective, this is a Gamma Poisson model. The Poisson distribution models patient counts and a Gamma distribution restricts the rate parameter of the Poisson to a reasonable range.

Selecting a prior distribution requires a lot of work. It helps, I believe to express the strength of the prior distribution as a fraction of the planned sample size. A prior with a strength equivalent to 50% of the planned sample size represents a setting you give equal weight to the prior distribution and the data when you have collected half of the planned sample size. That's a pretty strong prior, but not unreasonable if you are working on a clinical trial that is similar to many trials that you have run in the past.

## Boxplot of prior distribution

Try setting a distribution on the monthly accrual rate that has a mean equal to the expected accrual rate  but with varying levels of precision equivalent to `r paste0(round(100*s_list), "%")` of the planned sample size. 

### Figure 3.1. Boxplot of prior distribution for accrual rate.

```{r gamma_prior, fig.width=7, fig.height=1}
S <- 0.1
y_label1 <- "Monthly accrual rate (prior, S=0.1)"
sim1 <- data.frame(x=" ", y=rgamma(10000, N*S, T*S)*30)
S <- 0.2
y_label2 <- "Monthly accrual rate (prior, S=0.2)"
sim2 <- data.frame(x=" ", y=rgamma(10000, N*S, T*S)*30)
S <- 0.5
y_label3 <- "Monthly accrual rate (prior, S=0.5)"
sim3 <- data.frame(x=" ", y=rgamma(10000, N*S, T*S)*30)

y_range <- range(sim1$y, sim2$y, sim3$y)
custom_boxplot(sim1, y_label1, 1, "gray") + expand_limits(y=y_range)
custom_boxplot(sim2, y_label2, 1, "gray") + expand_limits(y=y_range)
custom_boxplot(sim3, y_label3, 1)         + expand_limits(y=y_range)

S <- 0.5
```

As a researcher who has done this sort of clinical trial many times, the value of S=`r S` seems reasonable to you. Smaller values of S produce distributions that are much wider than what you'd expect in a trial like this. With this value of S, you are ready to simulate your clinical trial.


Here is code in Stan (running inside of R) that simulates the results of a clinical trial with a target goal of 350 patients in 3 years (1095 days) and places a range of uncertainty on the accrual rate that is characterized by a gamma(175, 547.5). This is a prior distribution with a strength roughly equal to half the target sample size.

```{r gamma-poisson-before-trial, fig.width=7, fig.height=1}
f <- "3-gamma-poisson.stan"
# set n, t to zero before the trial starts
fit_gp1 <- stan(file=f,
  data=dat_prior, iter= 10000, warmup=1000, chains = 4)
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")

fit_gp1                                         %>%
  as.data.frame                                 -> sim_gp1
```

### Figure 3.2. Simulated total sample size prior to start of clinical trial.

```{r gamma-poisson-boxplot1, fig.width=7, fig.height=1}
# Boxplot of total sample size
y_label <- "Estimated total sample size (prior)"
sim_gp1                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(y_label, 1)                    -> prior_estimate
prior_estimate
```

You can learn a lot by looking at how the estimated total sample size relates to the prior paramter(s). 

Recall that in this simulation, you randomly select an accrual rate, gamma, and then you simulate a path based on that value of lambda. Not too surprisingly, the sample size depends on what lambda is chosen, though there is some variation, even for the same value of lambda.

### Figure 3.3. Relationship of prior estimate of accrual rate and estimated total sample size.

```{r gamma-poisson-scatterplot1, fig.width=7, fig.height=4}
# Scatterplot of accrual rate versus total sample size
sim_gp1                                         %>%
  use_series(lambda)                            %>%
  multiply_by(30)                               %>%
  quantile(pctl_list/100)                       -> x_ticks
x_ticks                                         %>%
  round(1)                                      -> x_labels
sim_gp1                                         %>%
  use_series(Nstar)                             %>%
  quantile(pctl_list/100)                       -> y_ticks
y_ticks                                         %>%
  round(1)                                      -> y_labels
sim_gp1                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=30*lambda)                           %>%
  ggplot(aes(x, y))                              +
  geom_point()                                   +
  ylab("Estimated total sample size (prior)")    + 
  xlab("Monthly accrual rate (prior)")           +
  scale_x_continuous(breaks=x_ticks,
                     minor_breaks=NULL,
                     labels=x_labels)            +
  scale_y_continuous(breaks=y_ticks,
                     minor_breaks=NULL,
                     labels=y_labels)
```

With a plot like this, you can run some sensitivity checks, such as "what would happen if the accrual rate were closer to 8.5 patients per month"

### Figure 3.4. Relationship of fixed accrual estimate and estimated total sample size.

```{r gamma-poisson-scatterplot2, fig.width=7, fig.height=4}
# Scatterplot of accrual rate versus total sample size
sim_gp1                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=30*lambda)                           %>%
  mutate(h=(x>8.45)*(x<8.55))                   %>%
  mutate(x=ifelse(h==0, x, 8.5))                %>%
  arrange(h)                                    -> sim_sliced
sim_sliced                                      %>%
  filter(h==1)                                  %>%  
  use_series(y)                                 %>%
  quantile(pctl_list/100)                       -> y_ticks
y_ticks                                         %>%
  round                                         -> y_labels
sim_sliced                                      %>%
  ggplot(aes(x, y))                              +
  ylab("Estimated total sample size (prior)")    + 
  xlab("Accrual rate (prior)")                   +
  geom_point(aes(color=factor(h)))               +    
  theme(legend.position="none")                  +
  scale_x_continuous(breaks=8.5,
                     minor_breaks=NULL)          +
  scale_y_continuous(breaks=y_ticks,
                     minor_breaks=NULL,
                     labels=y_labels)            +
  scale_color_manual(values=c("gray90", "red"))
```

## Update

The key reason that you should run your simulations in Stan is that you can make a seamless transition to a simulation of a clinical trial during the trial itself. In this trial, the early accrual rate was much lower than expected. After 239 days, you have only gotten 41 patients. If you were on target, you would have 239*(350/1095) = `r round(239/1095*350)` patients by now. How much is this shortfall hurting us?

### Figure 3.5. Simulated total sample size during the clinical trial.

```{r gamma-poisson-during-trial, fig.width=7, fig.height=1}
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.5, n=41, t=239)
fit_gp2 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)

fit_gp2              %>%
  as.data.frame      -> sim_gp2
y_range <- range(sim_gp1$Nstar, sim_gp2$Nstar)

# Boxplot of total sample size
prior_estimate + expand_limits(y=y_range)
y_label <- "Estimated total sample size (updated)"
sim_gp2                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(y_label, 1)                     +
  expand_limits(y=y_range)
```

## Post mortem analysis

For this trial, you were able to get 341 patients, but it took a lot longer than you expected, 1336 days instead of 1095 days. Time to get ready for your next clinical trial. NOT SO FAST! You are not done with simulations when the trial is over. After the trial, take a look at how your accrual rate ranks relative to the range of accrual rates associated with your prior distribution.

### Figure 3.6. Comparison of final accrual rate to prior distribution.

```{r gamma-poisson-after-trial, echo=FALSE, fig.width=7, fig.height=1}
sim4 <- data.frame(x=" ", y=30*341/1336)

sim3$y                                   %>%
  is_less_than(sim4$y)                   %>%
  mean                                   %>%
  multiply_by(100)                       %>%
  round(1)                               %>%
  paste0(" percentile")                  -> pctl
y_label4 <- paste0("Monthly accrual rate (final, ", pctl, ")")

y_range <- range(sim3$y, sim4$y)

# Boxplot of accrual rate
custom_boxplot(sim3, y_label3, 1) + expand_limits(y=y_range)
custom_boxplot(sim4, y_label4, 1) + expand_limits(y=y_range)
```

Okay, so you were way off. No one is going to hang you. If the same thing occurs over and over again, though, let's talk.

```{r save-everything}
save.image("3-gamma-poisson.RData")
```
