---
title: "Simulating clinical trials"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
```

## Hedging priors

One problem with informative prior distributions is that many researchers get them wrong. They are either too optimistic about the accrual rate, or they think they know the accrual rate with more certainty than they actually do, or sometimes both.

You can minimize the problems associated with a bad informative prior by adding a hedging hyperparameter.

```{r hedging-extension}
# hedging posterior
f <- "5-hedging-extension.stan"
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")
dat_during <- list(N=350, T=3*365, S=0.5, n=41, t=239)
fit_hd2 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_hd2                                         %>%
  as.data.frame                                 %>%
  mutate(i="hedging update")                    -> sim_hd2
```

```{r recalculate-informative-prior}
# informative prior
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.5, n=0, t=0)
fit_gp1 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_gp1                                         %>%
  as.data.frame                                 %>%
  mutate(i="prior estimate")                    -> sim_gp1
```

```{r recalculate-informative-update}
# informative posterior
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.5, n=41, t=239)
fit_gp2 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_gp2                                         %>%
  as.data.frame                                 %>%
  mutate(i="informative update")             -> sim_gp2
```

```{r calculate-flat-update}
# flat posterior
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.001/350, n=41, t=239)
fit_fl2 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_fl2                                         %>%
  as.data.frame                                 %>%
  mutate(i="flat update") -> sim_fl2
```

```{r calculate-data-projection}
# data
sim_dat <- data.frame(lambda=41/239, Nstar=41+(1095-239)*(41/239), i="data projection")
```

```{r hedging-boxplots, fig.width=5, fig.height=1.67}
# Boxplot of total sample size
sim_hd2                                         %>%
  bind_rows(sim_gp1)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

Notice that the large discrepancy with a pure data projection versus the informative prior. When there is a discrepancy, the hedging hyperprior will strongly downweight the informative prior.

The following boxplot shows how much the hedging hyperparameter downweighted the prior distribution

```{r downweight, fig.height=1, fig.width=5}
sim_hd2 %>%
  mutate(i=" ")                                 %>%
  ggplot(aes(i, pi))                             +
  expand_limits(y=c(0,2))                        +
  ylab("Hedging hyperparameter")                 + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

If you compared this to a flat prior, you would see that the flat prior would line up perfectly with the data projection.

```{r compare-to-flat, fig.width=7, fig.height=1.67}
# Boxplot of total sample size
sim_fl2                                         %>%
  bind_rows(sim_gp1)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

A non-Bayesian solution would probably also line up perfectly with the data projection.

```{r compare-to-informative, fig.width=7, fig.height=1.67}
sim_gp2                                         %>%
  bind_rows(sim_gp1)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

Let's look at the behavior of the hedging hyperprior when it's not needed--when the trial is actually slightly ahead of schedule (41 patients in 129 days).

```{r hedging-extension2}
# hedging posterior
f <- "5-hedging-extension.stan"
dat_during <- list(N=350, T=3*365, S=0.5, n=41, t=129)
fit_hd4 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_hd4                                         %>%
  as.data.frame                                 %>%
  mutate(i="hedging posterior")                 -> sim_hd4
```

```{r informative-prior2}
# informative prior
f <- "3-gamma-poisson.stan"
dat_before <- list(N=350, T=3*365, S=0.5, n=0, t=0)
fit_gp4 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_gp4                                         %>%
  as.data.frame                                 %>%
  mutate(i="informative prior")                 -> sim_gp4
```

```{r informative-update2}
# informative posterior
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.5, n=41, t=129)
fit_gp5 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_gp5                                         %>%
  as.data.frame                                 %>%
  mutate(i="informative posterior")             -> sim_gp5
```

```{r flat-update2}
# flat posterior
f <- "3-gamma-poisson.stan"
dat_during <- list(N=350, T=3*365, S=0.001/350, n=41, t=129)
fit_fl3 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
fit_fl3                                         %>%
  as.data.frame                                 %>%
  mutate(i="flat posterior")                    -> sim_fl3
```

```{r data-projection2}
# data
sim_dat <- data.frame(lambda=41/239, Nstar=41+(1095-129)*(41/129), i="Data")
```

```{r hedging-boxplot2, fig.width=7, fig.height=1.67}
# Boxplot of total sample size
sim_hd4                                         %>%
  bind_rows(sim_gp4)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

```{r hedging-boxplot3, fig.width=7, fig.height=1.67}
# Boxplot of total sample size
sim_fl3                                         %>%
  bind_rows(sim_gp4)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

```{r hedging-boxplot4, fig.width=7, fig.height=1.67}
# Boxplot of total sample size
sim_gp5                                         %>%
  bind_rows(sim_gp4)                            %>%
  bind_rows(sim_dat)                            %>%
  ggplot(aes(i, Nstar))                          +
  expand_limits(y=0)                             +
  ylab("Estimated total sample size")            + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```
The following boxplot shows how much the hedging hyperparameter downweighted the prior distribution. In this case, the data actually upweighted the prior slightly on average. Think of it as a reward for choosing your prior distribution so well. 

There's a technical fix that you can make so that the mean of the hyperparameter never exceeds 1, but we won't talk about it here.

```{r downweight2, fig.height=1, fig.width=7}
sim_hd4 %>%
  mutate(i=" ")                          %>%
  ggplot(aes(i, pi))                      +
  expand_limits(y=c(0,2))                 +
  ylab("Hedging hyperparameter")          + 
  xlab(" ")                               +
  geom_boxplot()                          +
  coord_flip()
```

