---
title: "Simulating clinical trials--beta extension"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r delay-preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
delay_priors <- list(
  N=N, T=T, S=S, 
  D1=D1, S1=S1,
  D2=D2, S2=S2,
  D3=D3, S3=S3,
  t=0, n=0)

delay_update1 <- delay_priors
delay_update1$t <- 150
delay_update1$n <- rep(0, 150)

pts <- unlist(read.csv("2-paths-count.csv"))
n2 <- sum(pts)
t2 <- 192 + length(pts)
delay_update2 <- delay_update1
delay_update2$t <- 192 + length(pts)
delay_update2$n <- c(rep(0, 192), pts)

save(N, T, S, S1, S2, S3, D1, D2, D3, t1, t2, n2, file="fig/6.0.RData")

N_label0 <- "Estimated total sample size (prior)"
N_label1 <- sub("prior", "update 1", N_label0)
N_label2 <- sub("prior", "update 2", N_label0)

d1_label0 <- "Days before start (prior)"
d1_label1 <- sub("prior", "update 1", d1_label0)
d1_label2 <- sub("prior", "update 2", d1_label0)

d2_label0 <- "Monthly accrual rate during warm-up phase (prior)"
d2_label1 <- sub("prior", "update 1", d2_label0)
d2_label2 <- sub("prior", "update 2", d2_label0)

d3_label0 <- "Days during warm-up phase (prior)"
d3_label1 <- sub("prior", "update 1", d3_label0)
d3_label2 <- sub("prior", "update 2", d3_label0)

l_label0 <- "Monthly accrual rate during late phase (prior)"
l_label1 <- sub("prior", "update 1", l_label0)
l_label2 <- sub("prior", "update 2", l_label0)
```

```{r setup-graphical-framework}
data.frame(x=c(100, 200), y=c(100, 250))             %>%
  ggplot(aes(x, y))                                   +
  geom_point(color="white")                           +
  labs(x=NULL, y=NULL)                                +
  geom_segment(x   =100, y   =150, 
               xend=120, yend=150,
               color="red", size=3)                   +
  geom_segment(x   =100, y   =150, 
               xend=120, yend=150,
               color="gray", size=1)                  +
  geom_label  (x=110, y=125, fill="red", 
               label="Waiting\nto start")             +
  geom_label  (x=135, y=125, fill="yellow",
               label="Early\naccrual")                 +
  geom_segment(x   =150, y   =200,
               xend=200, yend=200,
               color="green", size=3,
               arrow=arrow(length=unit(0.1, "inch"))) +
  geom_segment(x   =150, y   =200,
               xend=200, yend=200,
               color="gray", size=1, alpha=0.5,
               arrow=arrow(length=unit(0.1, "inch"))) +
  geom_label  (x=160, y=125, fill="green",
               label="Late\naccrual")                 + 
  theme(plot.margin=unit(c(0,0,0,0),"in"))            +
  theme(axis.line =element_blank())                   +
  theme(axis.text =element_blank())                   +
  theme(axis.ticks=element_blank())                   +
  theme(axis.title=element_blank())                  -> delay_graph
```

```{r delay-example1-prelims}
delay_graph                                      +
  geom_segment(x   =120, y   =180,
               xend=150, yend=180,
               color="yellow", size=3)           +
  geom_segment(x   =120, y   =180,
               xend=150, yend=180,
               color="gray", size=1)            -> fig
save(fig, file="fig/6.1.RData")
```

### Figure 6.1. Discontinuous transition from slow to normal accrual.
 
```{r delay-example1, fig.height=2, fig.width=5}
load("fig/6.1.RData"); print(fig)
```

```{r delay-example2-prelims}
delay_graph                                      +
  geom_segment(x   =120, y   =150,
               xend=150, yend=200,
               color="yellow", size=3)           +
  geom_segment(x   =120, y   =150,
               xend=150, yend=200,
               color="gray", size=1)            -> fig
save(fig, file="fig/6.2.RData")
```

### Figure 6.2. Linear transition from slow to normal accrual.

```{r delay-example2, fig.height=2, fig.width=5}
load("fig/6.2.RData"); print(fig)
```

```{r delay-example3-prelims}
delay_graph                                      +
  geom_segment(x   =120, y   =250,
               xend=150, yend=200,
               color="yellow", size=3)           +
    geom_segment(x   =120, y   =250,
               xend=150, yend=200,
               color="gray", size=1)            -> fig
save(fig, file="fig/6.3.RData")
```

### Figure 6.3. Linear transition from fast to normal accrual.

```{r delay-example3, fig.height=2, fig.width=5}
load("fig/6.3.RData"); print(fig)
```

```{r delay-example4-prelims}
x_original <- seq(0, 1, length=50)
y_original <- 3*x_original^2-2*x_original^3
x_rescaled <- 120 + 30 * x_original
y_rescaled <- 150 + 50 * y_original
df2 <- data.frame(x=x_rescaled, y=y_rescaled)
delay_graph                                      +
  geom_path(aes(x, y), data=df2,
            color="yellow", size=3)              +
  geom_path(aes(x, y), data=df2,
            color="gray",  size=1)              -> fig
save(fig, file="fig/6.4.RData")
```

### Figure 6.4. Smooth transition from slow to normal accrual.

```{r delay-example4, fig.height=2, fig.width=5}
load("fig/6.4.RData"); print(fig)
```

```{r delay-priors1-prelims}
d1_label0a <- sub("prior", paste("prior, S1", S1/2, sep="="), d1_label0)
d1_label0b <- sub("prior", paste("prior, S1", S1,   sep="="), d1_label0)
d1_label0c <- sub("prior", paste("prior, S1", S1*2, sep="="), d1_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S1/2*N*D1, S1/2*N*(1-D1))*T)
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S1  *N*D1, S1  *N*(1-D1))*T)
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S1*2*N*D1, S1*2*N*(1-D1))*T)

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplot(sim1, d1_label0a, 0, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.5a.RData")
custom_boxplot(sim2, d1_label0b, 0, color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.5b.RData")
custom_boxplot(sim3, d1_label0c, 0, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.5c.RData")
```

### Figure 6.5. Prior distribution for delta1 (time waiting to start).

```{r delay-priors1, fig.height=1, fig.width=10}
load("fig/6.5a.RData"); print(fig)
load("fig/6.5b.RData"); print(fig)
load("fig/6.5c.RData"); print(fig)
```

```{r delay-priors2-prelims}
d2_label0a <- sub("prior", paste("prior, S2", S2/2, sep="="), d2_label0)
d2_label0b <- sub("prior", paste("prior, S2", S2,   sep="="), d2_label0)
d2_label0c <- sub("prior", paste("prior, S2", S2*2, sep="="), d2_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S2/2*N*D2, S2/2*N*(1-D2))*30*N/T)
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S2  *N*D2, S2  *N*(1-D2))*30*N/T)
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S2*2*N*D2, S2*2*N*(1-D2))*30*N/T)

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplot(sim1, d2_label0a, 1, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.6a.RData")
custom_boxplot(sim2, d2_label0b, 1, color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.6b.RData")
custom_boxplot(sim3, d2_label0c, 1, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.6c.RData")
```

### Figure 6.6. Prior distribution for delta2 (accrual rate early), rescaled to patients per month.

```{r delay-priors2, fig.height=1, fig.width=10}
load("fig/6.6a.RData"); print(fig)
load("fig/6.6b.RData"); print(fig)
load("fig/6.6c.RData"); print(fig)
```

```{r delay-priors3-prelims}
d3_label0a <- sub("prior", paste("prior, S3", S3/2, sep="="), d3_label0)
d3_label0b <- sub("prior", paste("prior, S3", S3,   sep="="), d3_label0)
d3_label0c <- sub("prior", paste("prior, S3", S3*2, sep="="), d3_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S3/2*N*D3, S3/2*N*(1-D3))*T)
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S3  *N*D3, S3  *N*(1-D3))*T)
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S3*2*N*D3, S3*2*N*(1-D3))*T)

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplot(sim1, d3_label0a, 0, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.7a.RData")
custom_boxplot(sim2, d3_label0b, 0, color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.7b.RData")
custom_boxplot(sim3, d3_label0c, 0, color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/6.7c.RData")
```

### Figure 6.7. Prior distribution for delta3 (duration of early accrual), rescaled to days.

```{r delay-priors3, fig.height=1, fig.width=10}
load("fig/6.7a.RData"); print(fig)
load("fig/6.7b.RData"); print(fig)
load("fig/6.7c.RData"); print(fig)
```

### Here's the code to fit a simulation involving delays.

```{r delay_code}
f <- "6-beta-test3.jags"
cat(readLines(f), sep="\n")
```

```{r delay-fit0-prelims}
mod_dm0 <- jags.model(f, data=delay_priors, 
  n.adapt = 100)

parms <- c("delta1", "delta2", "delta3", "lambda", "Nstar")
fit_dm0 <- coda.samples(mod_dm0, var=parms, n.iter = 1000)
summary(fit_dm0)
fit_dm0                                         %>%
  as.matrix                                     %>%
  as.data.frame                                 -> sim_dm0
sim_dm0                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(N_label0, 0)                   -> fig
save(fig, file="fig/6.8.RData")
prior_estimate <- fig
```

### Figure 6.8. Prior estimate of total sample size, accounting for delays.

```{r delay-fit0, fig.height=1, fig.width=10}
load("fig/6.8.RData"); print(fig)
```

```{r delay-scaterplots-prelims}
sim_dm0                                         %>%
  mutate(x=delta1*T)                            %>%
  mutate(y=Nstar)                               %>%
  custom_scatterplot(d1_label0, N_label0, round_x=0) -> fig
save(fig, file="fig/6.9a.RData")
sim_dm0                                         %>%
  mutate(x=30*delta2*N/T)                       %>%
  mutate(y=Nstar)                               %>%
  custom_scatterplot(d2_label0, N_label0)       -> fig
save(fig, file="fig/6.9b.RData")
sim_dm0                                         %>%
  mutate(x=delta3*T)                            %>%
  mutate(y=Nstar)                               %>%
  custom_scatterplot(d3_label0, N_label0, round_x=0) -> fig
save(fig, file="fig/6.9c.RData")
sim_dm0                                         %>%
  mutate(x=lambda*30)                           %>%
  mutate(y=Nstar)                               %>%
  custom_scatterplot(l_label0, N_label0, round_x=0) -> fig
save(fig, file="fig/6.9d.RData")
```

### Figure 6.9. Effect of prior parameters on estimated total sample size.

```{r delay-scaterplots, fig.width=4, fig.height=4}
load("fig/6.9a.RData"); print(fig)
load("fig/6.9b.RData"); print(fig)
load("fig/6.9c.RData"); print(fig)
load("fig/6.9d.RData"); print(fig)
```

```{r delay-fit1-prelims-1}
mod_dm1 <- jags.model(f, data=delay_update1, 
  n.adapt = 100)

parms <- c("delta1", "delta2", "delta3", "lambda", "Nstar")
fit_dm1 <- coda.samples(mod_dm1, var=parms, n.iter = 1000)
```

```{r delay-fit1-prelims-2}
y_label <- "Estimated total sample size (update 1)"
f <- "6-beta-test3.jags"

fit_dm1                                         %>%
  as.matrix                                     %>%
  as.data.frame                                 -> sim_dm1
sim_dm1                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(N_label1, 0, co=color1)       -> update1
sim_dm0                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(N_label0, 0, co=color0)       -> prior_estimate
r <- range(sim_dm0$Nstar, sim_dm1$Nstar)
prior_estimate + expand_limits(y=r)             -> fig
save(fig, file="fig/6.10a.Rmd")
update1 + expand_limits(y=r)                    -> fig
save(fig, file="fig/6.10b.Rmd")
```

### Figure 6.10. Updated estimate of total sample size after 150 days of nothing.

```{r delay-fit1, fig.height=1, fig.width=10}
load("fig/6.10a.Rmd"); print(fig)
load("fig/6.10b.Rmd"); print(fig)
```

```{r delay-updated_start-prelims}
sim_dm0                                         %>%
  mutate(y=delta1*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d1_label0, 0, co=color0)      -> delta10
sim_dm1                                         %>%
  mutate(y=delta1*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d1_label1, 0, co=color1)      -> delta11
r <- T*range(sim_dm0$delta1, sim_dm1$delta1)
delta10 + expand_limits(y=r)                    -> fig
save(fig, file="fig/6.11a.RData")
delta11 + expand_limits(y=r)                    -> fig
save(fig, file="fig/6.11b.RData")
```

### Figure 6.11. How much longer do you expect to wait after 150 days of nothing.

```{r delay-updated_start, fig.width=10, fig.height=1}
load("fig/6.11a.RData"); print(fig)
load("fig/6.11b.RData"); print(fig)
```

```{r delay-update-delta2-prelims}
sim_dm0                                         %>%
  mutate(y=delta2*30*N/T)                       %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d2_label0, 1, co=color0)       -> delta20
sim_dm1                                         %>%
  mutate(y=delta2*30*N/T)                       %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d2_label1, 1, co=color1)       -> delta21
r <- range(sim_dm0$delta2, sim_dm1$delta2)*30*N/T
delta20 + expand_limits(y=r)                    -> fig
save(fig, file="fig/6.12a.RData")
delta21 + expand_limits(y=r)                    -> fig
save(fig, file="fig/6.12b.RData")
```

### Figure 6.12. Update of delta2 after 150 days of nothing.

```{r delay-update-delta2, fig.width=10, fig.height=1}
load("fig/6.12a.RData"); print(fig)
load("fig/6.12b.RData"); print(fig)
```

```{r delay-update-delta3-prelims}
sim_dm0                                         %>%
  mutate(y=delta3*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d3_label0, 0, co=color0)       -> delta30
sim_dm1                                         %>%
  mutate(y=delta3*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d3_label1, 0, co=color1)       -> delta31
r <- T*range(sim_dm0$delta3, sim_dm1$delta3)
delta30 + expand_limits(y=r) -> fig
save(fig, file="fig/6.13a.RData")
delta31 + expand_limits(y=r) -> fig
save(fig, file="fig/6.13b.RData")
```

### Figure 6.13. Update of delta3 after 150 days of nothing.

```{r delay-update-delta3, fig.width=10, fig.height=1}
load("fig/6.13a.RData"); print(fig)
load("fig/6.13b.RData"); print(fig)
```

```{r delay-update-lambda-prelims}
sim_dm0                                         %>%
  mutate(y=lambda*30)                           %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(l_label0, 1, co=color0)        -> lambda0
sim_dm1                                         %>%
  mutate(y=lambda*30)                           %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(l_label1, 1, co=color1)        -> lambda1
r <- range(sim_dm0$lambda, sim_dm1$lambda)*30
lambda0 + expand_limits(y=r) -> fig
save(fig, file="fig/6.14a.RData")
lambda1 + expand_limits(y=r) -> fig
save(fig, file="fig/6.14b.RData")
```

### Figure 6.14. Update of lambda after 150 days of nothing.

```{r delay-update-lambda, fig.width=10, fig.height=1}
load("fig/6.14a.RData"); print(fig)
load("fig/6.14b.RData"); print(fig)
```

### Update 2. We have 240 days of accrual data after 192 days of waiting.

```{r delay-fit2-prelims}
y_label <- "Estimated total sample size (update 2)"
f <- "6-beta-test3.jags"

mod_dm2 <- jags.model(f, data=delay_update2, 
  n.adapt = 100)

parms <- c("delta1", "delta2", "delta3", "lambda", "Nstar")
fit_dm2 <- coda.samples(mod_dm2, var=parms, n.iter = 1000)
summary(fit_dm2)
fit_dm2                                         %>%
  as.matrix                                     %>%
  as.data.frame                                 -> sim_dm2

r <- range(sim_dm0$Nstar, sim_dm2$Nstar)

sim_dm0                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(N_label0, 0, co=color0)         +
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.15a.RData")

sim_dm2                                         %>%
  rename(y=Nstar)                               %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(N_label2, 0, co=color1)         +
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.15b.RData")
```

### Figure 6.15. Second update of total sample size.

```{r delay-fit2, fig.height=1, fig.width=10}
load("fig/6.15a.RData"); print(fig)
load("fig/6.15b.RData"); print(fig)
```

```{r delay-update2-delta2-prelims}
r <- range(sim_dm0$delta2, sim_dm1$delta2, sim_dm2$delta2)*30*N/T
sim_dm0                                         %>%
  mutate(y=delta2*30*N/T)                       %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d2_label0, 1, co=color0)        + 
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.16a.RData")-> delta20

sim_dm2                                         %>%
  mutate(y=delta2*30*N/T)                       %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d2_label2, 1, co=color1)        + 
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.16b.RData")
```

### Figure 6.16. Second update of delta2.

```{r delay-update2-delta2, fig.width=10, fig.height=1}
load("fig/6.16a.RData"); print(fig)
load("fig/6.16b.RData"); print(fig)
```

```{r delay-update2-delta3-prelims}
r <- T*range(sim_dm0$delta3, sim_dm1$delta3, sim_dm2$delta3)
sim_dm0                                         %>%
  mutate(y=delta3*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d3_label0, 0, co=color0)        + 
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.17a.RData")

sim_dm2                                         %>%
  mutate(y=delta3*T)                            %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(d3_label2, 0)                   +
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.17b.RData")
```

### Figure 6.17. Second update of delta3.

```{r delay-update2-delta3, fig.width=10, fig.height=1}
load("fig/6.17a.RData"); print(fig)
load("fig/6.17b.RData"); print(fig)
```

```{r delay-update2-lambda-prelims}
r <- range(sim_dm0$lambda, sim_dm1$lambda, sim_dm1$lambda)*30
sim_dm0                                         %>%
  mutate(y=lambda*30)                           %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(l_label0, 1, co=color0)         +
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.18a.RData")

sim_dm2                                         %>%
  mutate(y=lambda*30)                           %>%
  mutate(x=" ")                                 %>%
  custom_boxplot(l_label2, 1, co=color1)         +
  expand_limits(y=r)                            -> fig
save(fig, file="fig/6.18b.RData")
```

### Figure 6.18. Second update of lambda.

```{r delay-update2-lambda, fig.width=10, fig.height=1}
load("fig/6.18a.RData"); print(fig)
load("fig/6.18b.RData"); print(fig)
```

```{r save_everything, echo=FALSE}
save.image("6-beta-extension.RData")
```
