---
title: "Simulating clinical trials--binomial extension"
author: "Steve Simon"
date: "March 22, 2017"
output:
  word_document: default
  html_document: default
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
P1 <- 0.1
S1 <- 0.4
P2 <- 0.3
S2 <- 0.2
n1 <- 90
n2 <- 81
n3 <- 41
save(N, T, S, P1, P2, S1, S2, n1, n2, n3, file="fig/7.0.RData")
dat_prior <- list(N=N, T=T, S=S, 
                   n=0, t=0, 
                   P1=P1, S1=S1, P2=P2, S2=S2,
                   n1=0, n2=0, n3=0)
dat_update <- list(N=N, T=T, S=S, 
                   P1=P1, S1=S1, P2=P2, S2=S2,
                   t=239, n1=90, n2=81, n3=41)
b1_label0 <- "Proportion excluded (prior)"
b1_label1 <- sub("prior", "update", b1_label0)
b2_label0 <- "Proportion refusing (prior)"
b2_label1 <- sub("prior", "update", b1_label0)
N_label1a <- "Estimated sample size before exclusions and refusals (prior)"
N_label2a <- "Estimated sample size after exclusions, before refusals (prior)"
N_label3a <- "Estimated sample size after exclusions and refusals (prior)"
N_label1b <- sub("prior", "update", N_label1a)
N_label2b <- sub("prior", "update", N_label2a)
N_label3b <- sub("prior", "update", N_label3a)
```

### Part 7. Modeling exclusions and refusals

```{r binomial-priors1-prelims}
b1_label0a <- sub("prior", paste("prior, S1", S1/4, sep="="), b1_label0)
b1_label0b <- sub("prior", paste("prior, S1", S1/2, sep="="), b1_label0)
b1_label0c <- sub("prior", paste("prior, S1", S1  , sep="="), b1_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S1/4*N*P1, S1/4*N*(1-P1)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S1/2*N*P1, S1/2*N*(1-P1)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S1  *N*P1, S1  *N*(1-P1)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplus(sim1, b1_label0a, 2, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1a.RData")
custom_boxplus(sim2, b1_label0b, 2, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1b.RData")
custom_boxplus(sim3, b1_label0c, 2, co=color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1c.RData")
```

### Figure 7.1. Prior distribution for pi1, proportion excluded.

```{r binomial-priors1, fig.height=1, fig.width=10}
load("fig/7.1a.RData"); print(fig)
load("fig/7.1b.RData"); print(fig)
load("fig/7.1c.RData"); print(fig)
```

```{r binomial-priors2-prelims}
b2_label0a <- sub("prior", paste("prior, S2", S2/4, sep="="), b2_label0)
b2_label0b <- sub("prior", paste("prior, S2", S2/2, sep="="), b2_label0)
b2_label0c <- sub("prior", paste("prior, S2", S2 , sep="="), b2_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S2/4*N*P2, S2/4*N*(1-P2)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S2/2*N*P2, S2/2*N*(1-P2)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S2  *N*P2, S2  *N*(1-P2)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplus(sim1, b1_label0a, 2, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2a.RData")
custom_boxplus(sim2, b1_label0b, 2, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2b.RData")
custom_boxplus(sim3, b1_label0c, 2, co=color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2c.RData")
```

### Figure 7.2. Prior distribution for pi2, proportion refusing.

```{r binomial-priors2, fig.height=1, fig.width=10}
load("fig/7.2a.RData"); print(fig)
load("fig/7.2b.RData"); print(fig)
load("fig/7.2c.RData"); print(fig)
```

### Here's what the code would look like for this model.

```{r binomial-code}
f <- "7-binomial-extension.stan"
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")
fit_bi0 <- stan(file=f,
  data=dat_prior, iter= 1000, chains = 4)
```

```{r binomial-prior-n-prelims}
fit_bi0                                         %>%
  as.data.frame                                 %>%
  mutate(x=" ")                                 -> sim_bi0

sim_bi0 %>% rename(y=N1star) -> sim_bi1
sim_bi0 %>% rename(y=N2star) -> sim_bi2
sim_bi0 %>% rename(y=N3star) -> sim_bi3
y_range <- range(sim_bi1$y, sim_bi2$y, sim_bi3$y)
prior1_plot <- custom_boxplus(sim_bi1, N_label1a, 0, color0)
prior2_plot <- custom_boxplus(sim_bi2, N_label2a, 0, color0)
prior3_plot <- custom_boxplus(sim_bi3, N_label3a, 0, color1)
prior1_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3a.RData")
prior2_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3b.RData")
prior3_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3c.RData")
```

### Figure 7.3. Prior estimate of sample size, accounting for exclusions and refusals.

```{r binomial-prior-n, fig.height=1, fig.width=10}
load("fig/7.3a.RData"); print(fig)
load("fig/7.3b.RData"); print(fig)
load("fig/7.3c.RData"); print(fig)
```

```{r binomial-scatterplots-prelims}
ylb <- "Estimated total sample size (prior)"
xlb <- "Proportion excluded (prior)"
sim_bi0                                         %>%
  select(pi1, N3star)                           %>%
  rename(x=pi1)                                 %>%
  rename(y=N3star)                              %>%
  custom_scatterplot(xlb, ylb, 2, 0)            -> fig
save(fig, file="fig/7.4a.RData")
xlb <- "Proportion refusing (prior)"
sim_bi0                                         %>%
  select(pi2, N3star)                           %>%
  rename(x=pi2)                                 %>%
  rename(y=N3star)                              %>%
  custom_scatterplot(xlb, ylb, 2, 0)            -> fig
save(fig, file="fig/7.4b.RData")
```

### Figure 7.4. How the prior distributions affect the estimated total sample size.

```{r binomial-scatterplots, fig.width=3, fig.height=3}
load("fig/7.4a.RData"); print(fig)
load("fig/7.4b.RData"); print(fig)
```

```{r binomial-update-n1-prelims}
f <- "7-binomial-extension.stan"
fit_bi4 <- stan(file=f,
  data=dat_update, iter= 1000, chains = 4)

fit_bi4                                         %>%
  as.data.frame                                 %>%
  mutate(x=" ")                                 -> sim_bi4

sim_bi4 %>% rename(y=N1star) -> sim_bi5
sim_bi4 %>% rename(y=N2star) -> sim_bi6
sim_bi4 %>% rename(y=N3star) -> sim_bi7
update2_plot <- custom_boxplus(sim_bi6, N_label2b, 0)
update3_plot <- custom_boxplus(sim_bi7, N_label3b, 0)
sim_bi1                                         %>%
  bind_rows(sim_bi2)                            %>%
  bind_rows(sim_bi3)                            %>%
  bind_rows(sim_bi5)                            %>%
  bind_rows(sim_bi6)                            %>%
  bind_rows(sim_bi7)                            %>%
  use_series(y)                                 %>%
  range                                         -> y_range

custom_boxplus(sim_bi1, N_label1a, 0, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.5a.RData")
custom_boxplus(sim_bi5, N_label1b, 0, co=color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.5b.RData")
```

### Figure 7.5. Updated sample size before exclusions and refusals.

```{r binomial-update-n1, fig.width=10, fig.height=1}
load("fig/7.5a.RData"); print(fig)
load("fig/7.5b.RData"); print(fig)
```

```{r binomial-update-n2-update}
custom_boxplus(sim_bi2, N_label2a, 0, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.6a.RData")
custom_boxplus(sim_bi6, N_label2b, 0, co=color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.6b.RData")
```

### Figure 7.6. Updated sample size after exclusions, before refusals.

```{r binomial-update-n2, fig.width=10, fig.height=1}
load("fig/7.6a.RData"); print(fig)
load("fig/7.6b.RData"); print(fig)
```

```{r binomial-update-n3-prelims}
custom_boxplus(sim_bi3, N_label3a, 0, co=color0) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.7a.RData")
custom_boxplus(sim_bi7, N_label3b, 0, co=color1) + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.7b.RData")
```

### Figure 7.7. Updated estimates after exclusions and refusals

```{r binomial-update-n3, fig.width=10, fig.height=1}
load("fig/7.7a.RData"); print(fig)
load("fig/7.7b.RData"); print(fig)
```

```{r save-everything}
save.image("7-binomial-extension.RData")
```
