---
title: "Simulating clinical trials--binomial extension"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
n_reps <- 1000
pctl_list <- c(1, 25, 50, 75, 99)
N <- 350
T <- 1095
S <- 0.5
P1 <- 0.1
S1 <- 0.1
P2 <- 0.3
S2 <- 0.2
n1 <- 90
n2 <- 81
n3 <- 41
save(N, T, S, P1, P2, S1, S2, n1, n2, n3, file="fig/7.0.RData")
dat_prior <- list(N=N, T=T, S=S, 
                   n=0, t=0, 
                   P1=P1, S1=S1, P2=P2, S2=S2,
                   n1=0, n2=0, n3=0)
dat_update <- list(N=N, T=T, S=S, 
                   P1=P1, S1=S1, P2=P2, S2=S2,
                   t=239, n1=90, n2=81, n3=41)
b1_label0 <- "Proportion excluded (prior)"
b1_label1 <- sub("prior", "update", b1_label0)
b2_label0 <- "Proportion refusing (prior)"
b2_label1 <- sub("prior", "update", b1_label0)
N_label1a <- "Estimated sample size before exclusions and refusals (prior)"
N_label2a <- "Estimated sample size after exclusions, before refusals (prior)"
N_label3a <- "Estimated sample size after exclusions and refusals (prior)"
N_label1b <- sub("prior", "update", N_label1a)
N_label2b <- sub("prior", "update", N_label2a)
N_label3b <- sub("prior", "update", N_label3a)
```

### Part 7. Modeling exclusions and refusals

There's a gauntlet that patients must travel through before they are officially part of your clinical trial. As you ask them questions, you will find that some of the patients might not meet all of your inclusion criteria. Also, the patients may change their minds, after hearing the details of your clinical trial, and decide that they don't really want to participate after all.

You can model these two processes using binomial distributions.

Here's some possible prior distributions for the proportion excluded and the proportion refusing consent.

```{r binomial-priors1-prelims}
b1_label0a <- sub("prior", paste("prior, S1", S1/2, sep="="), b1_label0)
b1_label0b <- sub("prior", paste("prior, S1", S1,   sep="="), b1_label0)
b1_label0c <- sub("prior", paste("prior, S1", S1*2, sep="="), b1_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S1/2*N*P1, S1/2*N*(1-P1)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S1  *N*P1, S1  *N*(1-P1)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S1*2*N*P1, S1*2*N*(1-P1)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplus(sim1, b1_label0a, 2, "gray") + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1a.RData")
custom_boxplus(sim2, b1_label0b, 2)         + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1b.RData")
custom_boxplus(sim3, b1_label0c, 2, "gray") + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.1c.RData")
```

### Figure 7.1. Prior distribution for pi1, proportion excluded.

```{r binomial-priors1, fig.height=1, fig.width=10}
load("fig/7.1a.RData"); print(fig)
load("fig/7.1b.RData"); print(fig)
load("fig/7.1c.RData"); print(fig)
```

A prior strength of 0.2 looks reasonable for the proportion excluded.

```{r binomial-priors2-prelims}
b2_label0a <- sub("prior", paste("prior, S2", S2/2, sep="="), b2_label0)
b2_label0b <- sub("prior", paste("prior, S2", S2,   sep="="), b2_label0)
b2_label0c <- sub("prior", paste("prior, S2", S2*2, sep="="), b2_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S2/2*N*P2, S2/2*N*(1-P2)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S2  *N*P2, S2  *N*(1-P2)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S2*2*N*P2, S2*2*N*(1-P2)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplus(sim1, b1_label0a, 2, "gray") + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2a.RData")
custom_boxplus(sim2, b1_label0b, 2)         + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2b.RData")
custom_boxplus(sim3, b1_label0c, 2, "gray") + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.2c.RData")
```

### Figure 7.2. Prior distribution for pi2, proportion refusing.

```{r binomial-priors2, fig.height=1, fig.width=10}
load("fig/7.2a.RData"); print(fig)
load("fig/7.2b.RData"); print(fig)
load("fig/7.2c.RData"); print(fig)
```

A prior strength of 0.1 looks reasonable for proportion refusing.

If you do a simple extrapolation, the total sample size after exclusions and refusals should be about `r 100*(1-P1)*(1-P2)`% of the `r N` planned patients or `r N*(1-P1)*(1-P2)` patients.

### Here's what the code would look like for this model.

```{r binomial-code}
f <- "7-binomial-extension.stan"
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")
fit_bi0 <- stan(file=f,
  data=dat_prior, iter= 1000, chains = 4)
```

```{r binomial-prior-n-prelims}
fit_bi0                                         %>%
  as.data.frame                                 %>%
  mutate(x=" ")                                 -> sim_bi0

sim_bi0 %>% rename(y=N1star) -> sim_bi1
sim_bi0 %>% rename(y=N2star) -> sim_bi2
sim_bi0 %>% rename(y=N3star) -> sim_bi3
y_range <- range(sim_bi1$y, sim_bi2$y, sim_bi3$y)
prior1_plot <- custom_boxplus(sim_bi1, N_label1a, 0)
prior2_plot <- custom_boxplus(sim_bi2, N_label2a, 0)
prior3_plot <- custom_boxplus(sim_bi3, N_label3a, 0)
prior1_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3a.RData")
prior2_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3b.RData")
prior3_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.3c.RData")
```

### Figure 7.3. Prior estimate of sample size, accounting for exclusions and refusals.

```{r binomial-prior-n, fig.height=1, fig.width=10}
load("fig/7.3a.RData"); print(fig)
load("fig/7.3b.RData"); print(fig)
load("fig/7.3c.RData"); print(fig)
```

You should consider exploring how sensitive your estimated total sample size is relative to each of the binomial parameters.

```{r binomial-scatterplots-prelims}
ylb <- "Estimated total sample size (prior)"
xlb <- "Proportion excluded (prior)"
sim_bi0                                         %>%
  select(pi1, N3star)                           %>%
  rename(x=pi1)                                 %>%
  rename(y=N3star)                              %>%
  custom_scatterplot(xlb, ylb, 2, 0)            -> fig
save(fig, file="fig/7.4a.RData")
xlb <- "Proportion refusing (prior)"
sim_bi0                                         %>%
  select(pi2, N3star)                           %>%
  rename(x=pi2)                                 %>%
  rename(y=N3star)                              %>%
  custom_scatterplot(xlb, ylb, 2, 0)            -> fig
save(fig, file="fig/7.4b.RData")
```

### Figure 7.4. How the prior distributions affect the estimated total sample size.

```{r binomial-scatterplots, fig.width=4, fig.height=4}
load("fig/7.4a.RData"); print(fig)
load("fig/7.4b.RData"); print(fig)
```

The estimated total sample size is more sensitive to the proportion consenting, mostly because your prior beliefs gave this proportion a broader range of possibilities.

### Update. 

The trial that you have seen illustrated thus far did not track exclusions and refusals, but let's suppose that there were `r n1` patients who showed up, but `r n1-n2` were excluded and `r n2-n3` who refused consent, leaving `r n3` available for randomization.

```{r binomial-update-n1-prelims}
f <- "7-binomial-extension.stan"
fit_bi4 <- stan(file=f,
  data=dat_update, iter= 1000, chains = 4)

fit_bi4                                         %>%
  as.data.frame                                 %>%
  mutate(x=" ")                                 -> sim_bi4

sim_bi4 %>% rename(y=N1star) -> sim_bi5
sim_bi4 %>% rename(y=N2star) -> sim_bi6
sim_bi4 %>% rename(y=N3star) -> sim_bi7
update1_plot <- custom_boxplus(sim_bi5, N_label1b, 0)
update2_plot <- custom_boxplus(sim_bi6, N_label2b, 0)
update3_plot <- custom_boxplus(sim_bi7, N_label3b, 0)
sim_bi1                                         %>%
  bind_rows(sim_bi2)                            %>%
  bind_rows(sim_bi3)                            %>%
  bind_rows(sim_bi5)                            %>%
  bind_rows(sim_bi6)                            %>%
  bind_rows(sim_bi7)                            %>%
  use_series(y)                                 %>%
  range                                         -> y_range
prior1_plot  + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.5a.RData")
update1_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.5b.RData")
```

### Figure 7.5. Updated sample size before exclusions and refusals.

```{r binomial-update-n1, fig.width=10, fig.height=1}
load("fig/7.5a.RData"); print(fig)
load("fig/7.5b.RData"); print(fig)
```

Your accrual rate is better than you expected, at least for getting people to show up.

```{r binomial-update-n2-update}
prior2_plot  + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.6a.RData")
update2_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.6b.RData")
```

### Figure 7.6. Updated sample size after exclusions, before refusals.

```{r binomial-update-n2, fig.width=10, fig.height=1}
load("fig/7.6a.RData"); print(fig)
load("fig/7.6b.RData"); print(fig)
```

The exclusions are about what you'd expect, and that means you're still ahead of schedule at this point.

```{r binomial-update-n3-prelims}
prior3_plot  + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.7a.RData")
update3_plot + expand_limits(y=y_range) -> fig
save(fig, file="fig/7.7b.RData")
```

### Figure 7.7. Updated estimates after exclusions and refusals

```{r binomial-update-n3, fig.width=10, fig.height=1}
load("fig/7.7a.RData"); print(fig)
load("fig/7.7b.RData"); print(fig)
```

The refusals, though, are a disaster. You are losing a lot more than `r 100*P2`% patients due to refusals. In fact, it looks like the surge in the proportion of refusals is the main factor influencing your behind-schedule numbers.

```{r save-everything}
save.image("7-binomial-extension.RData")
```
