---
title: "Simulating clinical trials--binomial extension"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
n_reps <- 1000
pctl_list <- c(1, 25, 50, 75, 99)
N <- 350
T <- 1095
S <- 0.5
P1 <- 0.1
S1 <- 0.1
P2 <- 0.3
S2 <- 0.2
b1_label0 <- "Proportion excluded (prior)"
b1_label1 <- sub("prior", "update", b1_label0)
b2_label0 <- "Proportion refusing (prior)"
b2_label1 <- sub("prior", "update", b1_label0)
```

### Part 7. Modeling exclusions and refusals

There's a gauntlet that patients must travel through before they are officially part of your clinical trial. As you ask them questions, you will find that some of the patients might not meet all of your inclusion criteria. Also, the patients may change their minds, after hearing the details of your clinical trial, and decide that they don't really want to participate after all.

You can model these two processes using binomial distributions.

Here's some possible prior distributions for the proportion excluded and the proportion refusing consent.

### Figure 7.1. Prior distribution for pi1, proportion excluded.

```{r binomial-priors1, fig.height=1, fig.width=7}
b1_label0a <- sub("prior", paste("prior, S1", S1/2, sep="="), b1_label0)
b1_label0b <- sub("prior", paste("prior, S1", S1,   sep="="), b1_label0)
b1_label0c <- sub("prior", paste("prior, S1", S1*2, sep="="), b1_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S1/2*N*P1, S1/2*N*(1-P1)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S1  *N*P1, S1  *N*(1-P1)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S1*2*N*P1, S1*2*N*(1-P1)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplot(sim1, b1_label0a, 2, "gray") + expand_limits(y=y_range)
custom_boxplot(sim2, b1_label0b, 2)         + expand_limits(y=y_range)
custom_boxplot(sim3, b1_label0c, 2, "gray") + expand_limits(y=y_range)
```

A prior strength of 0.2 looks reasonable for the proportion excluded.

### Figure 7.2. Prior distribution for pi2, proportion refusing.

```{r binomial-priors2, fig.height=1, fig.width=7}
b2_label0a <- sub("prior", paste("prior, S2", S2/2, sep="="), b2_label0)
b2_label0b <- sub("prior", paste("prior, S2", S2,   sep="="), b2_label0)
b2_label0c <- sub("prior", paste("prior, S2", S2*2, sep="="), b2_label0)

sim1 <- data.frame(x=" ", y=rbeta(n_reps, S2/2*N*P2, S2/2*N*(1-P2)))
sim2 <- data.frame(x=" ", y=rbeta(n_reps, S2  *N*P2, S2  *N*(1-P2)))
sim3 <- data.frame(x=" ", y=rbeta(n_reps, S2*2*N*P2, S2*2*N*(1-P2)))

y_range <- range(sim1$y, sim2$y, sim3$y)

custom_boxplot(sim1, b1_label0a, 2, "gray") + expand_limits(y=y_range)
custom_boxplot(sim2, b1_label0b, 2)         + expand_limits(y=y_range)
custom_boxplot(sim3, b1_label0c, 2, "gray") + expand_limits(y=y_range)
```

A prior strength of 0.1 looks reasonable for proportion refusing.

If you do a simple extrapolation, the total sample size after exclusions and refusals should be about `r 100*P1*P2`% of the 350 planned patients or `r N*P1*P2` patients.

### Here's what the code would look like for this model.

```{r binomial-code}
f <- "7-binomial-extension.stan"
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")
dat_before <- list(N=350, T=3*365, S=0.5, 
                   n=0, t=0, 
                   P1=0.9, S1=0.2, P2=0.7, S2=0.1,
                   n1=0, n2=0, n3=0)
fit_bi0 <- stan(file=f,
  data=dat_before, iter= 1000, chains = 4)
```

### Figure 7.3. Prior estimate of sample size, accounting for exclusions and refusals.

```{r binomial, fig.height=1, fig.width=7}
fit_bi0                                         %>%
  as.data.frame                                 %>%
  mutate(x="prior estimate") -> sim_bi0

y_label1 <- "Estimated total sample size"
y_label2 <- "Estimated sample size after exclusions"
y_label3 <- "Estimated sample size after exclusions and refusals"
sim_bi0 %>% rename(y=N1star) -> sim_bi1
sim_bi0 %>% rename(y=N2star) -> sim_bi2
sim_bi0 %>% rename(y=N3star) -> sim_bi3
y_range <- range(sim_bi1$y, sim_bi2$y, sim_bi3$y)
custom_boxplot(sim_bi1, y_label1, 0) + expand_limits(y=y_range)
custom_boxplot(sim_bi2, y_label2, 0) + expand_limits(y=y_range)
custom_boxplot(sim_bi3, y_label3, 0) + expand_limits(y=y_range)
```

You should consider exploring how sensitive your estimated total sample size is relative to each of the binomial parameters.

### Figure 7.4. How the prior distributions affect the estimated total sample size.

```{r binomial-part2, fig.width=4, fig.height=4}
sim_bi1                                         %>%
  ggplot(aes(pi1, N3star))                       +
  ylab("Estimated total sample size")            + 
  xlab("Proportion meeting inclusion")           +
  geom_point()

sim_bi1                                         %>%
  ggplot(aes(pi2, N3star))                       +
  ylab("Estimated total sample size")            + 
  xlab("Proportion consenting")                  +
  geom_point()

```

The estimated total sample size is more sensitive to the proportion consenting, mostly because your prior beliefs gave this proportion a broader range of possibilities.

### Update. 

The trial that you have seen illustrated thus far did not track exclusions and refusals, but let's suppose that there were `r n1` patients who showed up, but `r n1-n2` were excluded and `r n2-n3` who refused consent, leaving `r n3` available for randomization.

### Figure 7.5. Updated total sample size.

```{r binomial-extension-during, fig.width=7, fig.height=1}
f <- "7-binomial-extension.stan"
dat_before <- list(N=350, T=3*365, S=0.5, 
                   P1=0.9, S1=0.2, P2=0.7, S2=0.1,
                   t=239, n1=90, n2=81, n3=41)
fit_bi4 <- stan(file=f,
  data=dat_before, iter= 1000, chains = 4)

fit_bi4                                         %>%
  as.data.frame                                 %>%
  mutate(x="updated estimate")                  -> sim_bi4

y_label1 <- "Estimated total sample size"
y_label2 <- "Estimated sample size after exclusions"
y_label3 <- "Estimated sample size after exclusions and refusals"
sim_bi4 %>% rename(y=N1star) -> sim_bi5
sim_bi4 %>% rename(y=N2star) -> sim_bi6
sim_bi4 %>% rename(y=N3star) -> sim_bi7
y_range <- range(sim_bi1$y, sim_bi2$y, sim_bi3$y)
custom_boxplot(sim_bi5, y_label1, 0) + expand_limits(y=y_range)
custom_boxplot(sim_bi6, y_label2, 0) + expand_limits(y=y_range)
custom_boxplot(sim_bi7, y_label3, 0) + expand_limits(y=y_range)
```

The sample size estimate after refusals and exclusions is around 180 to 200, which is lower than what you were expecting prior to data collection, mostly because the proportion of refusals is a lot worse.

```{r save-everything}
save.image("7-binomial-extension.RData")
```
