---
title: "Simulating clinical trials"
author: "Steve Simon"
date: "March 22, 2017"
output: html_document
---

```{r preliminaries, echo=FALSE, message=FALSE}
source("0-preliminaries.R", echo=FALSE)
```

### Part 8. Modeling multi-center trials.

Here are the accrual data on two multi-center trials. The first, csp546, has accrual at 4 centers. The study ran for 56 weeks and recruited 613 total participants. The second, csp588, has accrual at 4 centers. The study ran for 24 weeks and recruited 127 patients.

```{r read-multicenter-data}

f <- "csp546 enrollment.csv"
csp546 <- read.csv(file=f, stringsAsFactors=FALSE, header=TRUE)

csp546a <- csp546
csp546a[is.na(csp546a)] <- 0
csp546a$Z <- apply(csp546a, 1, sum)
csp546b <- as.data.frame(sapply(csp546a, cumsum))
csp546b$t <- 1:(dim(csp546b)[1])
csp546b

f <- "csp558 enrollment.csv"
csp558 <- read.csv(file=f, stringsAsFactors=FALSE, header=TRUE)

csp558a <- csp558
csp558a[is.na(csp558a)] <- 0
csp558a$Z <- apply(csp558a, 1, sum)
csp558b <- as.data.frame(sapply(csp558a, cumsum))
csp558b$t <- 1:(dim(csp558b)[1])
csp558b
```

There are several ways to model accrual in a hierarchical setting. A simple and easy to follow approach is to model deviations of each individual center from the overall accrual rate with a lognormal distribution. The lognormal distribution is a bit tricky to work. One definition of a lognormal random variable is a random variable where the logarithm of that random variable is normally distributed. So you can describe the lognormal distribution in terms of the mean, $\mu$, and standard deviation, $\sigma$, of the underlying normal distribution. Or you can describe it in terms of the geometric mean, $e^\mu$, and geometric standard deviation, $e^\sigma$.

Another tricky issue is that the arithmetic mean of a lognormal random variable is going to be larger than the geometric mean. If X is lognormal with parameters $\mu$ and $\sigma$, then the arithmetic mean is larger than the geometric mean by a multiplicative factor, $e^{0.5~\sigma^2}$.

In our setting, we want to make sure that the arithmetic mean of the lognormal center effects is always 1. So our lognormals will use a "fudge factor" of $\mu=-0.5~\sigma^2$.

```{r set_constants}
n_centers <- 14
```

It's important to anchor these prior distributions to something concrete and observable. A simple numeric summary is the proportion of patients that are likely to come from your highest accruing center. Let's look at an example with `r n_centers` centers. If the accrual rates were identical across all `r n_centers` centers, then the each center would contribute roughly `r round(100/n_centers)`% of the patients.

### Figure 8.1. How much variation is produced by different lognormal distributions?

```{r examining-hierarchical-priors, fig.width=7, fig.height=2.67}
n_reps <- 1000

gsd_list <- c(1.1, 1.2, 1.5, 2, 4, 8)
n_gsd <- length(gsd_list)

lb <- rep(NA, n_gsd*n_reps)
max_pct <- rep(NA, n_gsd*n_reps)

j <- 0
for (gsd in gsd_list) {
  for (i in 1:n_reps) {
    j <- j+1
    eta <- rlnorm(n_centers, -0.5*log(gsd)^2, log(gsd))
    lb[j] <- paste("GSD =", gsd)
    max_pct[j] <- 100 * max(eta) / sum(eta)
  }
}
data.frame(lb, max_pct)                         %>%
  ggplot(aes(lb, max_pct))                       +
  expand_limits(y=c(0, 100))                     +
  ylab("Percent accrual in best center")         + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  geom_hline(yintercept=100/n_centers,
             color="gray")                       +
  coord_flip()
```

You decide after looking at these boxpots that a GSD around 1.5 is fairly representative of your experience with accrual in multi-center trials of this particular type and that a GSD of 2 represents the most extreme case you are likely to encounter.

Now you need to specify a prior on the GSD that is centered around 1.5 and where most of the probability is for values below 2.

### Figure 8.2. Prior distribution on the gsd.

```{r examining-hierarchical-priors-part2, fig.width=7, fig.height=2.67}
n_reps <- 1000
N <- 350

S1_list <- c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5)
n_S1 <- length(S1_list)

lb <- rep("---", n_S1*n_reps)
gsd <- rep(NA, n_S1*n_reps)

j <- 0
for (S1 in S1_list) {
  for (i in 1:n_reps) {
    j <- j+1
    gamma <- rgamma(1, N*S1, N*S1*log(1.5)^2)
    sigma <- sqrt(1 / gamma)
    gsd[j] <- exp(sigma)
    lb[j] <- paste("S1 =", S1)
  }
}
data.frame(lb, gsd)                             %>%
  ggplot(aes(lb, gsd))                           +
  ylab("GSD")                                    + 
  xlab(" ")                                      +
  geom_boxplot()                                 +
  coord_flip()
```

### Here's what the code looks like for a hierarchical model.

```{r hierarchical-code}
f <- "hierarchical-extension.stan"
# Here's what's hiding in the file 
cat(readLines(f), sep="\n")
dat_before <- list(N=350, T=3*365, S=0.5, M=14,
                   S1=0.02, GSD=1.5,
                   t=0, n=rep(0, 14))
fit_h21 <- stan(file=f,
  data=dat_before, iter= 1000, chains = 4)
```

### Figure 8.3. Estimated total sample size prior to start of clinical trial.

```{r hierarchical-boxplot, fig.width=7, fig.height=1}
fit_h21              %>%
  as.data.frame      %>%
  mutate(i="before") -> sim_h21
# Boxplot of total sample size
sim_h21                                  %>%
  ggplot(aes(i, Nstar))                   +
  expand_limits(y=0)                      +
  ylab("Estimated total sample size")     + 
  xlab(" ")                               +
  geom_boxplot()                          +
  coord_flip()
```

### Figure 8.4. Relationship of prior distributions to estimated total sample size.

```{r hierarchical-scatterplots}
sim_h21                                         %>%
  ggplot(aes(max_pct, Nstar))                    +
  ylab("Estimated total sample size")            + 
  xlab("Proportion accrued at best site")        +
  geom_point()

sim_h21                                         %>%
  ggplot(aes(gsd, max_pct))                      +
  xlab("GSD")                                    + 
  ylab("Proportion accrued at best site")        +
  geom_point()
```

```{r}
tidy(fit_h21, conf.int=TRUE)
m <- c(14, 9, 17, 0, 11, 28, 22, 1, 12, 25, 41, 0, 27, 0)	
length(m)
t <- 14
csp546b[t, 1:14]
f <- "hierarchical-extension.stan"
dat_during <- list(N=350, T=3*365, S=0.5, M=14,
                   S1=0.02, GSD=1.5,
                   t=14, n=csp546b[t, 1:14])
fit_h22 <- stan(file=f,
  data=dat_during, iter= 1000, chains = 4)
tidy(fit_h22)
```

```{r more-hierarchical}
fit_h21              %>%
  as.data.frame      %>%
  mutate(i="before") -> sim_h21

# Boxplot of total sample size
sim_h21                                  %>%
  ggplot(aes(i, Nstar))                   +
  expand_limits(y=0)                      +
  ylab("Estimated total sample size")     + 
  xlab(" ")                               +
  geom_boxplot()                          +
  coord_flip()
```

```{r more_stuff, eval=FALSE}
# Scatterplot of accrual rate versus total sample size
sim_h21                                  %>%
  ggplot(aes(30*mu, Nstar))               +
  ylab("Estimated total sample size")     + 
  xlab("Accrual rate")                    +
  geom_point()

sim_h21                                  %>%
  ggplot(aes(scale, Nstar))               +
  ylab("Estimated total sample size")     + 
  xlab("Hierarchical spread")             +
  geom_point()
```

Final comments???